Supplement to ``Machine Learning for Mechanistic Models of Metapopulation Dynamics''

### 2023-11-27 ms.Rnw
incorporating feedback from ERS

\begin{abstract}
Mathematical models in ecology and epidemiology must be consistent with observed data in order to generate reliable knowledge and sound policy. Metapopulation systems, which consist of a collection of sub-populations at various locations, pose technical challenges in statistical inference due to nonlinear, stochastic interactions. Difficulties encountered in these methodological issues can obstruct the core scientific questions concerning the link between the mathematical models and the data. Progress in statistically efficient simulation-based inference for partially observed stochastic dynamic systems has enabled the development of statistically rigorous approaches to the analysis of nonlinear but low-dimensional systems. Recently, an algorithm has been developed which enables comparable inference for higher-dimensional models arising in metapopulation systems. The COVID-19 pandemic provides a situation where mathematical models and their policy implications were widely visible, and we revisit an influential metapopulation model used to inform basic epidemiological understanding early in the pandemic. Our methods support self-critical data analysis, enabling us to identify and address model limitations, and leading to a new model with substantially improved statistical fit and parameter identifiability. Our results suggest that the lockdown initiated on January 23, 2020 in China was more effective than previously thought. We proceed to recommend statistical analysis standards for future metapopulation system modeling.
\end{abstract}

@@

A recent growth in the study of metapopulation dynamics has been driven partly by the COVID-19 pandemic \cite{li20,wu20,wang22,prieto22,cascante-vega22,pizzuti20,alleman21,yang21,engebretsen23} and in part by methodological advances facilitating the fitting of metapopulation models to spatiotemporal data.
Until the start of this millenium, developing dynamic models with both statistical and scientific justification was a longstanding open problem for even a single community \cite{bjornstad01}.
Over the past two decades, new algorithms \cite{ionides06,toni09,andrieu10,ionides15} and software \cite{king16,kristensen16,devalpine17} together with ever-increasing computational resources, have enabled routine inference for low-dimensional nonlinear partially observed stochastic dynamic systems.
However, fundamental algorithmic scalability issues known as the ``curse of dimensionality'' lead to difficulties with the high-dimensional systems arising in metapopulation inference.
These issues are clearest for Monte Carlo techniques based on importance sampling \cite{bengtsson08} but are also evident in the need for variational approximations for large Monte Carlo Markov Chain (MCMC) calculations \cite{blei17}. 
The quest for statistical methods suitable for metapopulation models of scientific interest has motivated the development of various methodological approaches.
Nevertheless, limitations have persisted in the capability to carry out flexible model-based inference and rigorous model criticism.
Thus, data analysis for metapopulation models has lagged behind low-dimensional time series analysis of biological systems.
Recent developments enable this gap to be closed, as we demonstrate via a reanalysis of COVID-19, viewed from the context of the ability to drawm evidence-based scientific conclusions about the dynamics of the emerging pandemic in January and February 2020.

@@

Biological systems are characterized by nonlinear stochastic dynamics together with incomplete and noisy measurements  \cite{bjornstad01}.
We therefore focus on the class of partially observed Markov process (POMP) models \cite{breto09}, acknowledging that deterministic models can be conceptually useful but are problematic as statistical explanations of noisy systems \cite{king15,wheeler23}.
The Markovian property asserts that the dynamic process has no memory conditional on its current state, which is algorithmically convenient while being scientifically nonrestrictive since we can choose what to include in the state.
Metapopulation models consider a multivariate system state at each location and so we require methods tailored for high-dimensional POMP models.
Simplifications arise if models and data are limited to binary presence-absence, or a small discrete set of values at each location \cite{mackenzie09}, but we focus on situations where abundance data are available, such as case reports for infectious diseases.

A natural place to look for statistical methodology applicable to metapopulation models is among the techniques developed for inferring population dynamics at a single location, reviewed by \cite{funk20,auger-methe21}.
Commonly implemented approaches for POMP models can be categorized as  (i) MCMC; (ii) matching summary statistics between data and simulations; (iii)  linearization;   (iv) particle filters (i.e., sequential Monte Carlo).

@@

Population dynamics may be approximately linear on a log scale, and this has been used to develop linearization methods for epidemiological time series analysis \cite{bjornstad02} that have been extended to metapopulation analysis \cite{xia04}.
This provides a numerically convenient set of tools, but requires scientists to work within a limited class of models. The ensemble Kalman filter (EnKF) developed in the context of massive geophysical model, combines an ensemble representation with a computationally efficient update rule inspired by the scalable linear Kalman filter, providing an approach with excellent scalability \cite{evensen09book,evensen22}.
For biological systems, EnKF was first demonstrated as a computationally convenient tool for compartment models at a single location \cite{shaman12,yang14}.
Subsequently, it has been applied for epidemiological metapopulation inference \cite{li20,kramer20-plos-cb}.
However, the linearization in the EnKF filter update rule can be problematic for highly nonlinear systems \cite{evensen22}.
Further, a linear update rule is not appropriate for small, discrete populations unless EnKF is embedded within a MCMC algorithm \cite{katzfuss19}.

For low-dimensional systems, particle filter methods are applicable to a flexible class of models for low-dimensional nonlinear systems \cite{breto09,king16}.
They have the attractive feature of statistically efficient use of data, i.e., accessing the likelihood function for Bayesian or likelihood-based inference, with approximation resulting only from finite Monte Carlo effort.
For high-dimensional systems, scalability considerations demand further approximations since particle filters suffer acutely from the ``curse of dimensionality'' \cite{bengtsson08}.
In contrast, block particle filter (BPF) methods achieve scalability by updating particles through localized resampling, rather than employing the linear update of EnKF \cite{rebeschini15}.
It is an empirical question which of these approximations is more successful on metapopulation models,  with prior evidence favoring BPF \cite{ionides23-jasa}.
In this paper, we not only aim to compare these methods in the context of metapopulation modeling but importantly also we reveal issues of model assessment when fitting complicated models to large datasets.
We introduce a model diagnosis procedure to mitigate the risk of selecting an inappropriate model, an essential step in informing public policy decisions.

@@

A standard way to convert a deterministic model, constructed as a system of ordinary differential equations, into a stochastic model is to reinterpret the rates of the deterministic system as rates of a Markov chain \citep{keeling09}.
This places limits on the mean-variance relationship of the resulting stochastic model \citep{breto11}.
Models allowing greater variability that permitted by this construction are said to be over-dispersed.
We added multiplicative white noise to the transmission rate, following the approach of \citep{breto09,he10}, giving rise to model~{\RevisedModelUnconstrained}.
We fitted the model using an iterated block particle filter to maximize the likelihood \citep{ionides22,ning23}.
Table~1 shows that this model outperforms simple statistical benchmarks, obtaining a competitive likelihood with relatively few parameters.
From a statistical perspective, {\RevisedModelUnconstrained} is therefore an adequate statistical description of the data.
However, some parameters of {\RevisedModelUnconstrained} were weakly identified by the data, especially in the pre-lockdown time interval within which there were relatively few reported cases (Supplementary Sec.~\suppSecUnconstrained).
When the evidence about the model parameters in the data is weak, the ambiguity may be resolved by other, unmodeled and poorly understood, aspects of the data.
This risks leading to undesirable situations where substantial conclusions about questions of interest could be driven by the weaknesses of the model rather than its strengths.

In Supplementary Sec.~\suppSecUnconstrained, we show how the flexibility of {\RevisedModelUnconstrained} can be used to obtain a high likelihood via an unplausibly long estimated duration of infection during the pre-lockdown period, with the estimate suddenly reducing after lockdown.
 {\RevisedModelConstrained} resolves this issue by constraining the latent and infectious periods to be the same before and after lockdown. 
We maximized the block particle filter likelihood using an iterated filtering algorithm \citep{ionides15} adapted to the structure of a block particle filter \citep{ionides22}.
The block filter approximation has also proven helpful for spatiotemporal inference when using alternatives to particle filtering and alternatives to maximization by iterated filtering \cite{whitehouse23}.
In the current context, the block particle filter was found to be more effective than a test suite of alternative filters including the ensemble Kalman filter (Supplementary Fig.~\suppFigFilterComparison).
Calculating the log-likelihood for each model in Table~\ref{tab:loglik} requires extensive computation to produce a single number which contains essentially all the information about the statistical fit of the model.
However, additional work is required to understand what characteristics of the models and data causes the differences in these numbers, and the practical consequences of the numerical results.

All parameter estimates are reported in Supplementary Table~\suppTableResults.
Here, we discuss the estimated basic reproductive number (i.e., the expected number of secondary infections from one index case in a fully susceptible population), denoted by $\Rzero^{\before}$ and $\Rzero^{\after}$ before and after the January 23rd lockdown.
The mathematical formula used to calculate  $\Rzero^{\before}$ and $\Rzero^{\after}$, in terms of the model parameters, is provided in Supplementary Table~\suppTableResults.
Our estimates for model {\RevisedModelConstrained}, are $\Rzero^{\before}=\Sexpr{metapoppkg::myround(ci_c["R0_be", "mle"],2)}$ with confidence interval (CI) (\Sexpr{metapoppkg::myround(ci_c["R0_be", "ci1"],2)},\Sexpr{metapoppkg::myround(ci_c["R0_be", "ci2"],2)}) and $\Rzero^{\after}=\Sexpr{metapoppkg::myround(ci_c["R0_af","mle"],2)}$  with CI (\Sexpr{metapoppkg::myround(ci_c["R0_af", "ci1"],2)},\Sexpr{metapoppkg::myround(ci_c["R0_af","ci2"],2)}), where the estimates and their associated 95\% CIs obtained by profile likelihood (Supplementary Sec.~\suppSecInference).
This implies that the Chinese government non-pharmaceutical interventions instituted on and around January 23 reduced $\Rzero$ by a factor of $\Sexpr{metapoppkg::myround(ci_c["R0_be","mle"]/ci_c["R0_af","mle"],1)}$.
By contrast, the estimates of \cite{li20} are $\widetilde\Rzero^{\before}=2.38$  with CI $(2.03,2.77)$ and $\widetilde\Rzero^{\after}=0.98$  with CI $(0.83,1.16)$, implying reduction by a factor of $\Sexpr{metapoppkg::myround(2.38/0.98,1)}$.
For comparison, interventions implemented across a panel of 41 countries (34 European) were estimated to reduce $\Rzero$ by a factor of $\Sexpr{metapoppkg::myround(1/(1-0.77),1)}$  with CI $(\Sexpr{metapoppkg::myround(1/(1-0.65),1)},\Sexpr{metapoppkg::myround(1/(1-0.85),1)})$ \citep{brauner21}.
Our estimate for $\Rzero$ before lockdown is toward the high end of previous estimates based on data up to February 2020, reviewed by \citep{park20-jcm}.
A contributing factor to our relatively high $\Rzero$ estimate may be that our mechanistic model includes transmission from asymptomatic individuals in the $\Rzero$ estimate \cite{subramanian21}.
An alternative metaopopulation analysis of the pre-lockdown China data, with a deterministic transmission model, obtained an $\Rzero$ estimate of 3.11  with CI  (2.39,4.13) \citep{read21}.

@@

A mechanistic model, whose likelihood is competitive with statistical benchmarks, is anticipated to have simulations that are qualitatively comparable to the data.
Since the model specification is inevitably imperfect, and is accounted for in the model fitting by noise processes, we expect simulations to have somewhat more stochastic variation than the data.
By contrast, models with insufficient flexibility to explain the data may give rise to simulations with too little stochasticity.
Comparing model {\LiParams} with {\RevisedModelUnconstrained} demonstrates this (Supplementary Fig.~\suppFigSim).
Fitted models having simulations with implausibly little variability lead to claims of excessive confidence about estimated parameters.
This may be clearest from a parametric bootstrap approach where CIs are constructed by re-estimating parameters using artificial datasets simulated from the fitted model.
However, it is also true for classical CI and Bayesian credible interval constructions.
Thus, CIs from mechanistic models, which outperform statistical benchmarks, are anticipated to be conservative, whereas CIs from models with insufficient variability to explain the data are generally anti-conservative.
Requiring model likelihoods to be comparable to statistical benchmarks therefore improves the credibility of uncertainty intervals as well as improving the accuracy of point estimates.

@@

Software environments are critical for providing data analysis that is not only reproducible but also readily extendable.
Scientists developing a data analysis should develop an environment that empowers them to explore their own models and data, and then they should share this environment as part of the publication process.
In practice, this involves encapsulating data analysis within a software package  that immerses the user in a documented environment where the models, methods and data used for the article can be readily be experimented with.
Trustworthy data analysis should be supported by unit testing and documentation, and the quality of this support should be one of the considerations in evaluation of the data analysis.
In other words, the article presenting the research should be part of a compendium \cite{gentleman07}.
The compendium for this article is comprised of the article source code, at \url{https://github.com/jifanli/metapop_article}, together with the software environment for the data analysis, provided by the R package at \url{https://github.com/jifanli/metapoppkg}.

@@

In conclusion, the study of metapopulation dynamics will continue to benefit from advances in algorithms, software and data analysis methodologies.
The models should undergo critical scrutiny to delineate their strengths and weaknesses, following evaluation procedures such as we have described in this paper. 
With due care, these models can unearth limitations in existing knowledge, investigate hypotheses that may extend our knowledge, and furnish us with valuable predictive tools.

@@

\noindent {\it Model}.
Our model is a spatiotemporal partially observed Markov process (SpatPOMP) \citep{ionides23-jasa}.
A partially observed Markov process (POMP) \citep{breto09,king16} is a stochastic dynamic model comprised of: (i) a latent process having the Markov property that the future is conditionally independent of the past given the present; (ii) a measurement process, describing how the data are modeled as noisy and incomplete observations of the latent process.
The SpatPOMP framework adds an extra assumption that the latent process is comprised of a collection of units, each of which has its own latent process and observation process.
The latent processes for each unit can be interdependent, but the observations for a given unit are required to depend only on the latent process for that unit.
In our context, each city is a unit.
General notation for SpatPOMP models is set up in Supplementary Sec.~\suppSecOurModel, and this notation is subsequently used to define mathematically the specific SpatPOMP models studied in this paper ({\LiMobility}, {\LiParams},{\RevisedModelUnconstrained} and {\RevisedModelConstrained}).
All the models under consideration have an SEAIR structure, as described in Figure~\ref{fig:flow_diagram}.

@@

\arxiv{}{
\section*{Code availability}
The code generating the manuscript is archived on Zenodo
\eic{ZENODO MS LINK HERE} and a development version is at \url{https://github.com:jifanli/metapop_article}.
The software environment developed for this analysis is also available as both an archived version (\eic{ZENODO METAPOPPKG LINK HERE}) and a development version (\url{https://github.com:jifanli/metapoppkg}).
These programs depend on other open-source software archived at \url{https://cran.r-project.org/}.
}

\section*{Acknowledgements}
This work was supported by National Science Foundation grants DMS-1761603 and DMS-1761612.
High performance research computing resources were provided by Texas A\&M University and University of Michigan.

>>>>

\begin{abstract}
Mathematical models in ecology and epidemiology must be consistent with observed data in order to generate reliable knowledge and evidence-based policy. Metapopulation systems, which consist of a collection of sub-populations at various locations, pose technical challenges in statistical inference due to nonlinear, stochastic interactions. Difficulties encountered in these methodological tasks can obstruct the core scientific questions concerning the link between the mathematical models and the data. Progress in statistically efficient simulation-based inference for partially observed stochastic dynamic systems has enabled the development of statistically rigorous approaches to the analysis of nonlinear but low-dimensional systems. Recently, an algorithm has been developed which enables comparable inference for higher-dimensional models arising in metapopulation systems. The COVID-19 pandemic provides a situation where mathematical models and their policy implications were widely visible, and we revisit an influential metapopulation model used to inform basic epidemiological understanding early in the pandemic. Our methods support self-critical data analysis, enabling us to identify and address model limitations, and leading to a new model with substantially improved statistical fit and parameter identifiability. Our results suggest that the lockdown initiated on January 23, 2020 in China was more effective than previously thought. We proceed to recommend statistical analysis standards for future metapopulation system modeling.
\end{abstract}

@@

A recent growth in the study of metapopulation dynamics has been driven partly by the COVID-19 pandemic \cite{li20,wu20,wang22,prieto22,cascante-vega22,pizzuti20,alleman21,yang21,engebretsen23} and in part by methodological advances facilitating the fitting of metapopulation models to spatiotemporal data.
Until the start of this millenium, developing dynamic models with both statistical and scientific justification was a longstanding open problem for even a single community \cite{bjornstad01}.
Over the past two decades, new algorithms \cite{ionides06,toni09,andrieu10,ionides15} and software \cite{king16,kristensen16,devalpine17}, together with ever-increasing computational resources, have enabled routine inference for low-dimensional nonlinear partially observed stochastic dynamic systems.
However, fundamental algorithmic scalability issues known as the ``curse of dimensionality'' lead to difficulties with the high-dimensional systems arising in metapopulation inference.
These issues are clearest for Monte Carlo techniques based on importance sampling \cite{bengtsson08} but are also evident in the need for variational approximations for large Monte Carlo Markov Chain (MCMC) calculations \cite{blei17}. 
The quest for statistical methods suitable for metapopulation models of scientific interest has motivated the development of various methodological approaches.
Nevertheless, limitations have persisted in the capability to carry out flexible model-based inference and rigorous model criticism.
Thus, data analysis for metapopulation models has lagged behind low-dimensional time series analysis of biological systems.
Recent developments enable this gap to be closed, as we demonstrate via a reanalysis of COVID-19, viewed from the context of the ability to drawm evidence-based scientific conclusions about the dynamics of the emerging pandemic in January and February 2020.

@@

Biological systems are characterized by nonlinear stochastic dynamics together with incomplete and noisy measurements  \cite{bjornstad01}.
We therefore focus on the class of partially observed Markov process (POMP) models \cite{breto09}, acknowledging that deterministic models can be conceptually useful but are problematic as statistical explanations of noisy systems \cite{king15,wheeler23}.
The Markov property asserts that the dynamic process has no memory conditional on its current state, which is algorithmically convenient while being scientifically nonrestrictive since we can choose what to include in the state.
Metapopulation models consider a multivariate system state at each location and so we require methods tailored for high-dimensional POMP models.
Simplifications arise if models and data are limited to binary presence-absence, or a small discrete set of values at each location \cite{mackenzie09}, but we focus on situations where abundance data are available, such as case reports for infectious diseases.

A natural place to look for statistical methodology applicable to metapopulation models is among the techniques developed for inferring population dynamics at a single location, reviewed by \cite{funk20,auger-methe21}.
Commonly implemented approaches for POMP models can be categorized as  (i) variants of MCMC; (ii) matching summary statistics between data and simulations; (iii)  linearization;   (iv) particle filters (i.e., sequential Monte Carlo).

@@

Population dynamics may be approximately linear on a log scale, and this has been used to develop linearization methods for epidemiological time series analysis \cite{bjornstad02} that have been extended to metapopulation analysis \cite{xia04}.
This provides a numerically convenient set of tools, but requires scientists to work within a limited class of models.

The ensemble Kalman filter (EnKF) developed in the context of massive geophysical model, combines an ensemble representation with a computationally efficient update rule inspired by the scalable linear Kalman filter, providing an approach with excellent scalability \cite{evensen09book,evensen22}.
For biological systems, EnKF was first demonstrated as a computationally convenient tool for compartment models at a single location \cite{shaman12,yang14}.
Subsequently, it has been applied for epidemiological metapopulation inference \cite{li20,kramer20-plos-cb}.
However, the linearization in the EnKF filter update rule can be problematic for highly nonlinear systems \cite{evensen22}.
Further, a linear update rule is not appropriate for small, discrete populations unless EnKF is embedded within a MCMC algorithm \cite{katzfuss19}.

For low-dimensional systems, particle filter methods are applicable to a flexible class of models: they permit consideration of arbitrary nonlinear dynamics and require the model to be specified only via a simulator \cite{breto09,king16}.
Particle filters enable statistically efficient use of data, since they provide an evaluation of the likelihood function required for Bayesian or likelihood-based inference, with approximation resulting only from finite Monte Carlo effort.
For high-dimensional systems, scalability considerations demand further approximations since particle filters suffer acutely from the ``curse of dimensionality'' \cite{bengtsson08}.
In contrast, block particle filter (BPF) methods achieve scalability by updating particles through localized resampling, rather than employing the linear update of EnKF \cite{rebeschini15}.
It is an empirical question which of these approximations is more successful on metapopulation models,  with prior evidence favoring BPF \cite{ionides23-jasa}.

Fitting complex models to large datasets using computationally intensive methods is the domain of machine learning.
Some machine learning models, such as artificial neural networks, lack a mechanistic interpretation.
Here, we focus instead on machine learning for mechanistic models.
We not only compare available machine learning methods for fitting metapopulation models but, importantly, we also consider issues of model criticism.
Specifically, we introduce a model diagnosis procedure to mitigate the risk of selecting an inappropriate model, an essential step in informing public policy decisions.

@@

A standard way to convert a deterministic model, constructed as a system of ordinary differential equations, into a stochastic model is to reinterpret the rates of the deterministic system as rates of a Markov chain \citep{keeling09}.
This places limits on the mean-variance relationship of the resulting stochastic model \citep{breto11}.
Models allowing greater variability that permitted by this construction are said to be over-dispersed.
We added multiplicative white noise to the transmission rate, following the approach of \citep{breto09,he10}, giving rise to model~{\RevisedModelUnconstrained}.
We fitted the model using an iterated block particle filter to maximize the likelihood \citep{ionides22,ning23}.
The block filter approximation has also proven helpful for spatiotemporal inference when using alternatives to particle filtering and alternatives to maximization by iterated filtering \cite{whitehouse23}.
In the current context, the block particle filter was found to be more effective for likelihood evaluation than a test suite of alternative filters including the ensemble Kalman filter (Supplementary Fig.~\suppFigFilterComparison).
The iterated block particle filter maximizes the block particle filter likelihood using an iterated filtering algorithm \citep{ionides15} adapted to the structure of a block particle filter.

Table~1 shows that model~{\RevisedModelUnconstrained} outperforms simple statistical benchmarks, obtaining a competitive likelihood with relatively few parameters.
From a statistical perspective, {\RevisedModelUnconstrained} is therefore an adequate statistical description of the data.
However, some parameters of {\RevisedModelUnconstrained} were weakly identified by the data, especially in the pre-lockdown time interval within which there were relatively few reported cases (Supplementary Sec.~\suppSecUnconstrained).
When the evidence about the model parameters in the data is weak, the ambiguity may be resolved by other, unmodeled and poorly understood, aspects of the data.
This risks leading to undesirable situations where substantial conclusions about questions of interest could be driven by the weaknesses of the model rather than its strengths.
In Supplementary Sec.~\suppSecUnconstrained, we show how the flexibility of {\RevisedModelUnconstrained} can be used to obtain a high likelihood via an unplausibly long estimated duration of infection during the pre-lockdown period, with the estimate suddenly reducing after lockdown.
 We resolved this issue by constraining the latent and infectious periods to be the same before and after lockdown, leading to model {\RevisedModelConstrained}.
 The additional constraints of {\RevisedModelConstrained} lead to a small loss of likelihood compared to {\RevisedModelUnconstrained}, but the fit remains competitive compared to the benchmark models, and the stronger identifiability facilitates the interpretation of estimated parameters.
Calculating the log-likelihood for each model in Table~\ref{tab:loglik} requires extensive computation to produce a single number which contains essentially all the information about the statistical fit of the model.
However, additional work is required to understand what characteristics of the models and data causes the differences in these numbers, and the practical consequences of the numerical results.

Parameter values for models~ {\LiMobility},~{\RevisedModelUnconstrained} and~{\RevisedModelConstrained} are reported in Supplementary Table~\suppTableResults.
Here, we discuss the estimated basic reproductive number (i.e., the expected number of secondary infections from one index case in a fully susceptible population), denoted by $\Rzero^{\before}$ and $\Rzero^{\after}$ before and after the January 23rd lockdown.
The mathematical formula used to calculate  $\Rzero^{\before}$ and $\Rzero^{\after}$, in terms of the model parameters, is provided in Supplementary Table~\suppTableResults.
Our estimates for model {\RevisedModelConstrained}, are $\Rzero^{\before}=\Sexpr{metapoppkg::myround(ci_c["R0_be", "mle"],2)}$ with confidence interval (CI) (\Sexpr{metapoppkg::myround(ci_c["R0_be", "ci1"],2)},\Sexpr{metapoppkg::myround(ci_c["R0_be", "ci2"],2)}) and $\Rzero^{\after}=\Sexpr{metapoppkg::myround(ci_c["R0_af","mle"],2)}$  with CI (\Sexpr{metapoppkg::myround(ci_c["R0_af", "ci1"],2)},\Sexpr{metapoppkg::myround(ci_c["R0_af","ci2"],2)}), where the estimates and their associated 95\% CIs obtained by profile likelihood (Supplementary Sec.~\suppSecInference).
This implies that the Chinese government non-pharmaceutical interventions instituted on and around January 23 reduced $\Rzero$ by a factor of $\Sexpr{metapoppkg::myround(ci_c["R0_be","mle"]/ci_c["R0_af","mle"],1)}$.
By contrast, the estimates of \cite{li20} are $\widetilde\Rzero^{\before}=2.38$  with CI $(2.03,2.77)$ and $\widetilde\Rzero^{\after}=0.98$  with CI $(0.83,1.16)$, implying reduction by a factor of $\Sexpr{metapoppkg::myround(2.38/0.98,1)}$.
For comparison, interventions implemented across a panel of 41 countries (34 European) were estimated to reduce $\Rzero$ by a factor of $\Sexpr{metapoppkg::myround(1/(1-0.77),1)}$  with CI $(\Sexpr{metapoppkg::myround(1/(1-0.65),1)},\Sexpr{metapoppkg::myround(1/(1-0.85),1)})$ \citep{brauner21}.
Our estimate for $\Rzero$ before lockdown is toward the high end of previous estimates based on data up to February 2020, reviewed by \citep{park20-jcm}.
An alternative metaopopulation analysis of the pre-lockdown China data, with a deterministic transmission model, obtained an $\Rzero$ estimate of 3.11  with CI  (2.39,4.13) \citep{read21}.
Our $\Rzero$ estimate is consistent with pre-lockdown estimates from other locations, such as New York city, for models that include asymptomatics \cite{subramanian21}.

@@

If a mechanistic model has likelihood competitive with statistical benchmarks, it is anticipated to have simulations that are qualitatively comparable to the data.
Since the model specification is inevitably imperfect, and is accounted for in the model fitting by noise processes, we expect simulations from the fitted model to have somewhat more stochastic variation than the data.
By contrast, models which are structurally unable to provide sufficient variability to explain the data must give rise to simulations with too little stochasticity.
Comparing model {\LiParams} with {\RevisedModelUnconstrained} demonstrates this (Supplementary Fig.~\suppFigSim).
Models that have simulations with implausibly little variability give rise to claims of excessive confidence about the uncertainty surrounding estimated parameters.
This phenomenon may be clearest when CIs are calculated using  parametric bootstrap approach, involving re-estimation of parameters using artificial datasets simulated from a fitted model.
However, it also applies for classical CI and Bayesian credible interval constructions.
Thus, CIs from mechanistic models that outperform statistical benchmarks are anticipated to be conservative, whereas CIs from models with insufficient variability to explain the data are generally anti-conservative.
Requiring model likelihoods to be comparable to statistical benchmarks therefore improves the credibility of uncertainty intervals as well as improving the accuracy of point estimates.

@@

Software environments are critical for providing data analysis that is not only reproducible but also readily extendable.
Scientists developing a data analysis should build an environment that empowers them to explore their own models and data, and then they should share this environment as part of the publication process.
In practice, this involves encapsulating data analysis within a software package  that immerses the user in a documented environment where the models, methods and data used for the article can be readily be experimented with.
Trustworthy data analysis should be supported by unit testing and documentation, and the quality of this support should be one of the considerations in evaluation of the data analysis.
In other words, the article presenting the research should be part of a compendium \cite{gentleman07}.
The compendium for this article is comprised of the article source code, at \url{https://github.com/jifanli/metapop_article}, together with the software environment for the data analysis, provided by the R package at \url{https://github.com/jifanli/metapoppkg}.

@@

In conclusion, the study of metapopulation dynamics will continue to benefit from advances in algorithms, software, and data analysis methodologies.
The models should undergo critical scrutiny to delineate their strengths and weaknesses, following evaluation procedures such as we have described in this paper. 
With due care, these models can unearth limitations in existing knowledge, investigate hypotheses that may extend our knowledge, and furnish us with valuable predictive tools.

@@

\noindent {\it Model}.
Our model is a spatiotemporal partially observed Markov process (SpatPOMP) \citep{ionides23-jasa}.
A partially observed Markov process (POMP) \citep{breto09,king16} is a stochastic dynamic model comprised of: (i) a latent process having the Markov property that the future is conditionally independent of the past given the present; (ii) a measurement process, describing how the data are modeled as noisy and incomplete observations of the latent process.
The SpatPOMP framework adds an extra assumption that the latent process is comprised of a collection of units, each of which has its own latent process and observation process.
The latent processes for each unit can be interdependent, but the observations for a given unit are required to depend only on the latent process for that unit.
In our context, each city is a unit.
General notation for SpatPOMP models is set up in Supplementary Sec.~\suppSecOurModel, and this notation is subsequently used to define mathematically the specific SpatPOMP models studied in this paper ({\LiMobility}, {\LiParams}, {\RevisedModelUnconstrained} and {\RevisedModelConstrained}).
All the models under consideration have an SEAIR structure, as described in Figure~\ref{fig:flow_diagram}.

@@

\arxiv{}{
\section*{Code availability}
The code generating the manuscript is archived on Zenodo (\url{https://zenodo.org/records/10149258}) and a development version is at \url{https://github.com:jifanli/metapop_article}.
The software environment developed for this analysis is also available as both an archived version (\url{https://zenodo.org/records/10149233}) and a development version (\url{https://github.com:jifanli/metapoppkg}).
These programs depend on other open-source software archived at \url{https://cran.r-project.org/}.
}

\section*{Acknowledgements}
This work was supported by National Science Foundation grants DMS-1761603 and DMS-1761612.
Portions of this research were conducted with Texas A\&M High Performance Research Computing and University of Michigan Advanced Research Computing.
We acknowledge discussions with Ethan Romero-Severson.

### 2023-11-27 si/si.Rnw
incorporating feedback from ERS

The performance of particle filters decays rapidly with the dimension of the latent state \citep{bengtsson08}.
Iterated particle filters suffer from the same curse of dimensiounality.
Block particle filters avoid this curse by partitioning the latent states into weakly dependent units, and resampling separately on each unit.
This is analogous to recombination in sexual reproduction, with each block corresponding to a chromosome.
The recombination allows successful blocks of one particle to join up with different successful blocks from another particle.
If this approximation permits effective high-dimensional filtering then the parameter perturbation strategy can be employed for parameter estimation, just as for basic particle filters.

>>>>

The performance of particle filters decays rapidly with the dimension of the latent state \citep{bengtsson08}.
Iterated particle filters suffer from the same curse of dimensionality.
Block particle filters avoid this curse by partitioning the latent states into weakly dependent units, and resampling separately on each unit.
This is analogous to recombination in sexual reproduction, with each block corresponding to a chromosome.
In this evolutionary analogy, each particle at time $t_n$ is an individual in the $n$th generation of a population.
The latent state of the particle is its genetic material, and this state is divided into chromosomes corresponding to each block.
Reproduction occurs at the resampling stage of the block particle filter, at which point the next generation of particles is resampled from the pool of chomosomes.
In this resampling process, each chromosome from the previous generation is selected proportionally to its fitness.
Thus, recombination allows successful blocks of one particle to join up with different successful blocks from another particle.
If this approximation permits effective high-dimensional filtering then the parameter perturbation strategy can be employed for parameter estimation, just as for basic particle filters.
